---
title: "AQI Demo"
author: Walter Wang, Sherry Zhang
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AQI Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Below we will introduce Air Quality Index (AQI), and demonstrate how to use tidyindex package to compute it based on existing datasets.

```{r setup, echo = TRUE, message = FALSE}
library(tidyindex)
library(dplyr)
library(purrr)
```

# 1. Everything about AQI

The Air Quality Index (AQI) is an index created to measure air quality, with high values indicating greater pollution levels. The index categorized into six levels, each corresponding to a specific numerical range: 

  * Good (0 - 50), 
  * Moderate (50 - 100), 
  * Unhealthy for Sensitive Groups (101 - 150),
  * Unhealthy (151 - 200), 
  * Very Unhealthy (201 - 300), and 
  * Hazardous (300 and higher). 

The AQI is calculated individually for each of the six major pollutants: Ozone (O3), PM2.5, PM10, Carbon Monoxide (CO), Sulfur Dioxide (SO2), Nitrogen Dioxide (NO2). The pollution concentrations are first truncated, and then the AQI value for each pollutant is calculated based on the following equation:

$$\text{I}_\text{p} = \frac{\text{I}_\text{Hi} - \text{I}_\text{Lo}}{\text{BP}_\text{Hi} - \text{BP}_\text{Lo}}(\text{C}_\text{p} - \text{BP}_\text{Lo}) + \text{I}_\text{Lo}.$$
where:

* $\text{I}_\text{p}$ is the AQI for pollutant $\text{p}$,

* $\text{C}_\text{p}$ is the truncated concentration of pollutant $\text{p}$,

* $\text{BP}_\text{Hi}$ and $\text{BP}_\text{Lo}$ are the high and low concentration breakpoints of $\text{C}_\text{p}$,

* $\text{I}_\text{Hi}$ and $\text{I}_\text{Lo}$ are the AQI values corresponding to $\text{BP}_\text{Hi}$ and $\text{BP}_\text{Lo}$, respectively.

Each individual AQI value is rounded to the nearest integer, and the highest value among the six pollutant is taken as the final AQI.

The breakpoint values for each pollutant are provided in Table 6 of the [Technical Assistance Document for the Reporting of Daily Air Quality](https://document.airnow.gov/technical-assistance-document-for-the-reporting-of-daily-air-quailty.pdf). 

As an example, with an 8-hour measurement of CO value of 4.67, this value is first truncate to 4.6. For CO concentrations ranging from 4.5 to 9.4, the corresponding AQI range is 51 to 100. Using the formula above, the AQI is calculated as follows:

$$\frac{100 - 51}{9.4 - 4.5}\times(4.6 - 4.5) + 51 = 75.01,$$
This gives an AQI value of 75 for CO after rounding.

# 2. Data

To obtain all AQI-related data, we use the aqsr package, which provides an R interface for the EPA Air Quality System (AQS) API. For details about how to use the package, see the [aqsr GitHub repository](https://github.com/kpkeller/aqsr). 

In this repository, we have provided some AQI data for convenience. `aqi_travis` contains the PM2.5 values in Travis county, Austin, Texas, USA from 2024/01/01 to 2024/03/31, measured at three monitor sites. `aqi_table` and `pollutant_table` contain breakpoint values for AQI and six major pollutants, respectively.

# 3. Tidyindex Demonstration 

In this section, we will demonstrate how to use tidyindex to compute AQI.

## 3.1 Load Datasets

First, let's load the dataset:

```{r}
aqi
```

## 3.2 Find Breakpoints

According to the equation given in section 1, to compute AQI, we need to find the high and low breakpoints for every sample measurement, as well as the corresponding AQI breakpoints. We can use the following helper functions to lookup the breakpoint information.

```{r}
lookup_helper <- function(sample, subset){
  return(subset %>% filter(sample >= low & sample <= high))
}

aqi_lookup <- function(dataset){
  # takes the dataset of measurements as input
  # returns a tibble object with corresponding breakpoints and group info
  id <- dataset$code[1]
  concentration <- dataset$value
  if(id == "44201"){
    concentration <- trunc(concentration * 10^3)/10^3
  }
  else if (id == "88101" | id == "42101"){
    concentration <- trunc(concentration * 10)/10
  }
  else{
    concentration <- trunc(concentration)
  }
  subset <- pollutant_ref_tbl %>% filter(code == id)
  results <- map_dfr(concentration, ~ lookup_helper(.x, subset)) %>%
    bind_cols(aqi |> select(-pollutant, -code))
  return(results)
}
```

Here, the `lookup_helper` function will find the high and low breakpoints for any given samples. The `aqi_lookup` function first truncates every measurement according to different types of pollutants, then calls `lookup_helper` to return a tibble object, which contains every sample measurement, their corresponding high/low breakpoints, and group information. 

```{r}
breakpoints <- aqi_lookup(aqi)
breakpoints
```

## 3.3 Initialize Tidyindex Pipeline

With all preparation done, let's now initialize the tidyindex pipeline. 

```{r}
pipeline <- init(breakpoints)
```

#### Note that to efficiently use the tidyindex pipeline, it is recommended to put all needed data columns into one single dataframe or tibble object, as we have in `breakpoints`.

Upon inspection of the AQI equation in section 1, we know that the computation of AQI could be decomposed into two parts: a min-max rescaling with $\text{min} = \text{BP}_\text{Lo}$ and $\text{max} = \text{BP}_\text{Hi}$, as well as a transformation, which multiplies $\text{I}_\text{Hi} - \text{I}_\text{Lo}$ and then adds $\text{I}_\text{Lo}$.

Therefore, we can easily construct the pipeline using the `rescaling()` and `variable_transformation()` modules provided in tidyindex. 

## 3.4 Rescaling

The minmax rescaling is done by calling the `rescale_minmax()` function.

```{r}
pipeline <- pipeline |> 
  rescaling(minmax = rescale_minmax(value, min = low, max = high))
```

The `rescale_minmax()` function supports vectorized data inputs, so here we can just leave `sample_measurements`, `low_breakpoint` and `high_breakpoint` as vectors from `breakpoints`, and the function will automatically perform the minmax rescaling across every single data entry. The rescaling results are stored in the `minmax` column added to the pipeline.

```{r}
pipeline$data %>% dplyr::select(pollutant, date, low, high, value, minmax)
```

## 3.5 Variable Transformation

Next step is variable transformation, where we wish to multiply and add to the original variable. Such transformation is called affine transformation. While it is not already written in `tidyindex`, one can always come up with and implement their customized transformation functions, as well as other necessary functions. For our case, the transformation could be implemented by adding the following lines:

```{r eval = FALSE}
#' @rdname variable-transformation
#' @export
trans_affine <- function(var, a = NULL, b = NULL){
  fn <- function(x, a = NULL, b = NULL) a*x + b
  new_trans("trans_affine", var = enquo(var), fn = fn, a = enquo(a), b = enquo(b))
}
```

Here, we name it as `trans_affine`, and it has three inputs: `var`, the variable that will be transformed; `a`, the multiplicative coefficient, and `b`, the additive constant. These inputs will be used in the next line to construct the transformation function `fn`, which has our desired form `a*x + b`. Then, we create the `trans_affine` function by calling `new_trans()`, which will register the name, as well as all other inputs including the variable itself, the transformation function `fn`, and parameters `a` and `b`.

#### Note: here we put `enquo()` around `a` and `b` to make sure they are being evaluated in quosure, so that we can find them in the dataframe. Otherwise we will have to manually create these variables and pass them to the function for evaluation.

With the implemented transformation, we can now finish our pipeline.

```{r}
pipeline <- pipeline |> 
  variable_trans(AQI = trans_affine(minmax, a = high - low, b = low))
pipeline$data$AQI <- round(pipeline$data$AQI)
pipeline$data |> dplyr::select(pollutant, date, low, high, value, minmax, AQI)
```

# 4. Visualization

Now let's check our results. Below we show a line graph with AQI values from all three monitor sites, computed using our pipeline.

```{r}
library(ggplot2)
pipeline$data$date <- as.Date(pipeline$data$date)
ggplot(pipeline$data, aes(x = date, y = AQI, color = site_name)) +
  geom_line() + 
  labs(title = "AQI Values Over Time by Site",
       x = "Date",
       y = "AQI",
       color = "Monitor Sites") +
  theme_minimal() +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  theme(legend.position = "bottom")
```



