---
title: "AQI Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AQI Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Below we will introduce Air Quality Index (AQI), and demonstrate how to use tidyindex package to compute it based on existing datasets.

```{r setup, echo = TRUE, message = FALSE}
library(tidyverse)
library(tidyindex)
library(aqsr)
library(dplyr)
library(purrr)
```

# 1. Everything about AQI

Air Quality Index (AQI) is an index created to measure air quality. A higher value of AQI indicates a greater level of pollution throughout a certain area. AQI is divided into six categories based on different range of values: Good (0 - 50), Moderate (50 - 100), Unhealthy for Sensitive Groups (101 - 150), Unhealthy (151 - 200), Very Unhealthy (201 - 300) and Hazardous (300 and higher). 

How is AQI calculated? Firstly, we measure concentrations of six major pollutants: Ozone (O3), PM2.5, PM10, Carbon Monoxide (CO), Sulfur Dioxide (SO2), Nitrogen Dioxide (NO2). Then we truncate each measurement according to the requirements and compute the AQI based on the equation below:

$$\text{I}_\text{p} = \frac{\text{I}_\text{Hi} - \text{I}_\text{Lo}}{\text{BP}_\text{Hi} - \text{BP}_\text{Lo}}(\text{C}_\text{p} - \text{BP}_\text{Lo}) + \text{I}_\text{Lo}.$$
where:

$\text{I}_\text{p}$ stands for the index for pollutant $\text{p}$,

$\text{C}_\text{p}$ stands for the truncated concentration of pollutant $\text{p}$,

$\text{BP}_\text{Hi}$ and $\text{BP}_\text{Lo}$ represent the high and low concentration breakpoints of $\text{C}_\text{p}$,

$\text{I}_\text{Hi}$ and $\text{I}_\text{Lo}$ represent the AQI values corresponding to $\text{BP}_\text{Hi}$ and $\text{BP}_\text{Lo}$, respectively.

The value is then rounded to the nearest integer.

For detailed information including the table of AQI breakpoints, see the [Technical Assistance Document for the Reporting of Daily Air Quality](https://document.airnow.gov/technical-assistance-document-for-the-reporting-of-daily-air-quailty.pdf). 

After calculating AQI values for all six pollutants, the maximum value among all six is taken as the final result.

For example, suppose we have an 8-hour measurement of CO value as 4.67. We first truncate it to 4.6, then find the high and low breakpoints as 4.5 and 9.4, respectively. The corresponding AQI values of 4.5 and 9.4 are 51 and 100, and hence we can compute the AQI value for the 8-hour CO measurement as follows:

$$\frac{100 - 51}{9.4 - 4.5}\times(4.6 - 4.5) + 51 = 75.01,$$
We round the value to its nearest integer, resulting in a final AQI of 75.

# 2. Data

To obtain all AQI-related data, we use the aqsr package, which provides an R interface for the EPA Air Quality System (AQS) API. For details about how to use the package, see the [aqsr GitHub repository](https://github.com/kpkeller/aqsr). 

In this repository, we have provided some AQI data for convenience. `aqi_travis` contains the PM2.5 values in Travis county, Austin, Texas, USA from `2024/01/01` to `2024/03/31`, measured from three monitor sites. `aqi_table` and `pollutant_table` contain breakpoint values for AQI and six major pollutants, respectively.

# 3. Tidyindex Demonstration 

In this section, we will demonstrate how to use tidyindex to compute AQI.

## 3.1 Load Datasets

First, let's load the dataset:

```{r}
aqi_dataset <- aqi_travis %>% dplyr::select(state, city, county, parameter, parameter_code, date_local, local_site_name, sample_measurement)

aqi_dataset
```

## 3.2 Find Breakpoints

According to the equation given in section 1, to compute AQI, we need to find the high and low breakpoints for every sample measurement, as well as the corresponding AQI breakpoints. We can use the following helper functions to lookup the breakpoint information.

```{r}
lookup_helper <- function(sample, subset){
  return(subset %>% filter(sample >= low_breakpoint & sample <= high_breakpoint))
}

aqi_lookup <- function(dataset){
  # takes the dataset of measurements as input
  # returns a tibble object with corresponding breakpoints and group info
  id <- dataset$parameter_code[1]
  concentration <- dataset$sample_measurement
  if(id == "44201"){
    concentration <- trunc(concentration * 10^3)/10^3
  }
  else if (id == "88101" | id == "42101"){
    concentration <- trunc(concentration * 10)/10
  }
  else{
    concentration <- trunc(concentration)
  }
  subset <- pollutant_table %>% filter(pollutant_code == id)
  results <- map_dfr(concentration, ~ lookup_helper(.x, subset)) %>% left_join(aqi_table, "group")
  results$date <- dataset$date_local
  results$local_site_name <- dataset$local_site_name
  results$sample_measurement <- concentration
  return(results)
}
```

Here, the `lookup_helper` function will find the high and low breakpoints for any given samples. The `aqi_lookup` function first truncates every measurement according to different types of pollutants, then calls `lookup_helper` to return a tibble object, which contains every sample measurement, their corresponding high/low breakpoints, and group information. 

```{r}
breakpoints <- aqi_lookup(aqi_dataset)
breakpoints
```

## 3.3 Initialize Tidyindex Pipeline

With all preparation done, let's now initialize the tidyindex pipeline. 

```{r}
pipeline <- init(breakpoints)
```

#### Note that to efficiently use the tidyindex pipeline, it is recommended to put all needed data columns into one single dataframe or tibble object, as we have in `breakpoints`.

Upon inspection of the AQI equation in section 1, we know that the computation of AQI could be decomposed into two parts: a min-max rescaling with $\text{min} = \text{BP}_\text{Lo}$ and $\text{max} = \text{BP}_\text{Hi}$, as well as a transformation, which multiplies $\text{I}_\text{Hi} - \text{I}_\text{Lo}$ and then adds $\text{I}_\text{Lo}$.

Therefore, we can easily construct the pipeline using the `rescaling()` and `variable_transformation()` modules provided in tidyindex. 

## 3.4 Rescaling

The minmax rescaling is done by calling the `rescale_minmax()` function.

```{r}
pipeline <- pipeline |> rescaling(minmax = rescale_minmax(sample_measurement,
                                       min=low_breakpoint,
                                       max=high_breakpoint))
```

The `rescale_minmax()` function supports vectorized data inputs, so here we can just leave `sample_measurements`, `low_breakpoint` and `high_breakpoint` as vectors from `breakpoints`, and the function will automatically perform the minmax rescaling across every single data entry. The rescaling results are stored in the `minmax` column added to the pipeline.

```{r}
pipeline$data %>% dplyr::select(minmax)
```

## 3.5 Variable Transformation

Next step is variable transformation, where we wish to multiply and add to the original variable. Such transformation is called affine transformation. While it is not already written in `tidyindex`, one can always come up with and implement their customized transformation functions, as well as other necessary functions. For our case, the transformation could be implemented by adding the following lines:

```{r eval = FALSE}
#' @rdname variable-transformation
#' @export
trans_affine <- function(var, a = NULL, b = NULL){
  fn <- function(x, a = NULL, b = NULL) a*x + b
  new_trans("trans_affine", var = enquo(var), fn = fn, a = enquo(a), b = enquo(b))
}
```

Here, we name it as `trans_affine`, and it has three inputs: `var`, the variable that will be transformed; `a`, the multiplicative coefficient, and `b`, the additive constant. These inputs will be used in the next line to construct the transformation function `fn`, which has our desired form `a*x + b`. Then, we create the `trans_affine` function by calling `new_trans()`, which will register the name, as well as all other inputs including the variable itself, the transformation function `fn`, and parameters `a` and `b`.

#### Note: here we put `enquo()` around `a` and `b` to make sure they are being evaluated in quosure, so that we can find them in the dataframe. Otherwise we will have to manually create these variables and pass them to the function for evaluation.

With the implemented transformation, we can now finish our pipeline.

```{r}
pipeline <- pipeline |> variable_trans(AQI = trans_affine(minmax, 
                                                          a=AQI_high_breakpoint - AQI_low_breakpoint, 
                                                          b=AQI_low_breakpoint))
pipeline$data$AQI <- round(pipeline$data$AQI)
pipeline$data %>% dplyr::select(AQI)
```

# 4. Visualization

Now let's check our results. Below we show a line graph with AQI values from all three monitor sites, computed using our pipeline.

```{r}
library(ggplot2)
pipeline$data$date <- as.Date(pipeline$data$date)
ggplot(pipeline$data, aes(x = date, y = AQI, color = local_site_name)) +
  geom_line() + 
  labs(title = "AQI Values Over Time by Site",
       x = "Date",
       y = "AQI",
       color = "Monitor Sites") +
  theme_minimal() +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



