---
title: "AQI Demo"
author: Walter Wang, Sherry Zhang
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AQI Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Below we will introduce Air Quality Index (AQI), and demonstrate how to use tidyindex package to compute it based on existing datasets.

```{r setup, echo = TRUE, message = FALSE}
library(tidyindex)
library(dplyr)
library(purrr)
```

# Everything about AQI

The Air Quality Index (AQI) is an index created to measure air quality, with high values indicating greater pollution levels. The index categorized into six levels, each corresponding to a specific numerical range: 

  * Good (0 - 50), 
  * Moderate (50 - 100), 
  * Unhealthy for Sensitive Groups (101 - 150),
  * Unhealthy (151 - 200), 
  * Very Unhealthy (201 - 300), and 
  * Hazardous (300 and higher). 

The AQI is calculated individually for each of the six major pollutants: Ozone (O3), PM2.5, PM10, Carbon Monoxide (CO), Sulfur Dioxide (SO2), Nitrogen Dioxide (NO2). The pollution concentrations are first truncated, and then the AQI value for each pollutant is calculated based on the following equation:

$$\text{I}_\text{p} = \frac{\text{I}_\text{Hi} - \text{I}_\text{Lo}}{\text{BP}_\text{Hi} - \text{BP}_\text{Lo}}(\text{C}_\text{p} - \text{BP}_\text{Lo}) + \text{I}_\text{Lo}.$$
where:

* $\text{I}_\text{p}$ is the AQI for pollutant $\text{p}$,

* $\text{C}_\text{p}$ is the truncated concentration of pollutant $\text{p}$,

* $\text{BP}_\text{Hi}$ and $\text{BP}_\text{Lo}$ are the high and low concentration breakpoints of $\text{C}_\text{p}$, which are provided in Table 6 of the [Technical Assistance Document for the Reporting of Daily Air Quality](https://document.airnow.gov/technical-assistance-document-for-the-reporting-of-daily-air-quailty.pdf). 

* $\text{I}_\text{Hi}$ and $\text{I}_\text{Lo}$ are the AQI values corresponding to $\text{BP}_\text{Hi}$ and $\text{BP}_\text{Lo}$, respectively.

Each individual AQI value is rounded to the nearest integer, and the highest value among the six pollutant is taken as the final AQI.

As an example, with an 8-hour measurement of CO value of 4.67, this value is first truncate to 4.6. For CO concentrations ranging from 4.5 to 9.4, the corresponding AQI range is 51 to 100. Using the formula above, the AQI is calculated as follows:

$$\frac{100 - 51}{9.4 - 4.5}\times(4.6 - 4.5) + 51 = 75.01,$$

This gives an AQI value of 75 for CO after rounding.

# Construct AQI with `tidyindex`

In this section, we will demonstrate how to compute the AQI using the `tidyindex` package. The `aqi` data, available in the `tidyindex` package, contains daily PM2.5 values recorded from January to March 2024 at three monitoring sites in Travis county, Texas, USA.

```{r}
aqi
```

## Augment low/high breakpoints

To compute the AQI, as outlined in Section 1, we must first determine the high and low breakpoints for each sample measurement, as well as the corresponding AQI breakpoints. The helper functions below are designed to assist with retrieving this breakpoint information and augmenting them into the `aqi` data.

```{r}
lookup_helper <- function(sample, subset){
  return(subset %>% filter(sample >= low & sample <= high))
}

breakpoint_lookup <- function(dataset){
  # takes the dataset of measurements as input
  # returns a tibble object with corresponding breakpoints and group info
  id <- dataset$code[1]
  concentration <- dataset$value
  if(id == "44201"){
    concentration <- trunc(concentration * 10^3)/10^3
  }
  else if (id == "88101" | id == "42101"){
    concentration <- trunc(concentration * 10)/10
  }
  else{
    concentration <- trunc(concentration)
  }
  subset <- pollutant_ref_tbl %>% filter(code == id)
  results <- map_dfr(concentration, ~ lookup_helper(.x, subset)) %>%
    bind_cols(aqi |> select(-pollutant, -code))
  return(results)
}
```

The `lookup_helper()` function finds the high and low breakpoints for any given samples. The `breakpoint_lookup()` function first truncates each measurement based on the pollutant, then calls `lookup_helper()` to augment the high/low breakpoints and group information.

```{r}
(aqi_aug <- breakpoint_lookup(aqi))
```

*Note that to efficiently use the tidyindex pipeline, it is recommended to put all needed data columns into one single dataframe or tibble object.*

The construction of AQI can be divided into two steps: 

  - a minmax **rescaling** step: $I_{rescaled} = \frac{\text{C}_\text{p} - \text{BP}_\text{Lo}}{\text{BP}_\text{Hi} - \text{BP}_\text{Lo}}$ 
  - a **variable transformation** step (affine transformation): $(\text{I}_\text{Hi} - \text{I}_\text{Lo}) \times I_{rescaled} + \text{I}_\text{Lo}$.

## Rescaling

The minmax rescaling is performed by the `rescaling` module and the `rescale_minmax()` step requires `min` and `max` argument, which correspond to teh  low and high breakpoints we've looked up.  The rescaled values are stored in the `minmax` column.

```{r}
pipeline <- init(aqi_aug) |> 
  rescaling(minmax = rescale_minmax(value, min = low, max = high))

pipeline$data %>% dplyr::select(pollutant, date, low, high, value, minmax)
```

## Variable transformation

While an affine transformation step is not initially included in the `tidyindex` package, you can always easily create one:  

```{r eval = FALSE}
trans_affine <- function(var, a = NULL, b = NULL){
  fn <- function(x, a = NULL, b = NULL) a*x + b
  new_trans("trans_affine", var = enquo(var), 
            fn = fn, a = enquo(a), b = enquo(b))
}
```

The `trans_affine` function has three inputs: 

  * `var`, the variable to be transformed; 
  * `a`, the multiplicative coefficient, and 
  * `b`, the additive constant
  
The `fn` function specifies the transformation as `a*x + b`. The `new_trans()` constructor is then called to register the name along with other inputs, including the transformed variable, the transformation function `fn`, and associated parameters `a` and `b`. This creates a "recipe" for the transformation that can be evaluated standalone and used in the `variable_trans()` module. 

```{r}
trans_affine(minmax, a = high - low, b = low)
```

*Note: the `enquo()` function is required for additional parameters`a` and `b` in the `new_trans()` constructor. This creates a defused expression that describes how to evaluate the values but not actually execute the evaluation. For more details, see `rlang::enquo()`*

This concludes our pipeline for computing the AQI: 

```{r}
pipeline <- pipeline |> 
  variable_trans(AQI = trans_affine(minmax, a = high - low, b = low))
pipeline$data$AQI <- round(pipeline$data$AQI)
pipeline$data |> dplyr::select(pollutant, date, low, high, value, minmax, AQI)
```

Now let's check our results. Below we show a line graph with AQI values from all three monitor sites, computed using our pipeline.

```{r}
library(ggplot2)
pipeline$data$date <- as.Date(pipeline$data$date)
ggplot(pipeline$data, aes(x = date, y = AQI, color = site_name)) +
  geom_line() + 
  labs(title = "AQI Values Over Time by Site",
       x = "Date",
       y = "AQI",
       color = "Monitor Sites") +
  theme_minimal() +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  theme(legend.position = "bottom")
```



